Deciding Bitvector Termination with Program Synthesis

### 2014-06-24 papers/termination/paper.tex
revised motivation
revised related work + motivation + preliminaries

 The main contributions of our work can be summarised as follows:
\begin{itemize}
\item We designed a technique for computing ranking functions that correctly accounts for the wrap-around behavior caused by under- and overflows in bit-vector and floating point arithmetic. To the best of our knowledge, this is the first approach able to compute ranking functions for programs handling floats. Our technique is not restricted to finding linear ranking functions, but can also compute (lexicographic) non-linear  ones. We justified the need for such a non restrictive procedure by computing the probability ... .
\item  We rephrased the termination problem as a second-order satisfaction problem and made 
use of results in genetic programming to efficiently solve it. We have also investigated the effects of genetic operators on the search space for ranking functions and computed theoretical 
bounds on the convergence time ...
\item We present a formulation that allows us to prove termination without an expensive reachability check (as we don't use Ramsey-based termination arguments).  In particular,
we only need a bounded model checker that performs a single unwinding of a loop.
%% \item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
%% multiple loops.
\item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
\end{itemize}

@@

@@

The program in Figure~\ref{fig:motivation.h} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
\begin{verbatim}
R(x, y, z) = (x < y, z)
\end{verbatim}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0 && y > 0 && z > 0) {
    if (y > x) {
      y = z;
      x = nondet();
      z = x - 1;
    } else {
      z = z - 1;
      x = nondet();
      y = x - 1;
    }
}
\end{lstlisting}
\caption{}
 \label{fig:motivation.h}
\end{subfigure}

@@

@@

\section{Preliminaries}
\subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
\todo{linear program}

A transition system with state space $X$ and transition relation $T \subseteq X \times X$
is said to be \emph{unconditionally terminating} if there is no infinite sequence
$x_1, x_2, \ldots$ with $\forall i . T(x_i, x_{i+1})$.  We can prove that $T$ is
unconditionally terminating by finding an injective function $R: X \to Y$ where
$Y$ is well-founded and $R$ is monotonically decreasing with respect $T$.  That is
to say:
$$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$

@@

\subsection{Disjunctive Well-Foundedness}\label{sec:dwf}
Terminator uses disjunctive well foundedness arguments, which can be found
piece by piece, but require reasoning about the transitive closure of a transition
relation.  Because we build a monolithic well foundedness argument, we
do not need to reason about the full transitive closure, and can instead do our
reasoning over a single step of the transition relation.  This means that we
can get away with very simple model checking methods, rather than requiring
a full-blown safety prover.

@@

The existence of a ranking function for a single loop is equivalent to the satisfiability
of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
$R$ is a ranking function that proves the loop $L$ terminates.

>>>>

 The main contributions of our work can be summarised as follows:
\begin{itemize}
\item We designed a technique for computing ranking functions that correctly accounts for the wrap-around behavior caused by under- and overflows in bit-vector and floating point arithmetic. To the best of our knowledge, this is the first approach able to compute ranking functions for programs handling floats. Our technique is not restricted to finding linear ranking functions, but can also compute (lexicographic) non-linear  ones. We justified the need for such a non restrictive procedure by computing the probability ... .
\item  We rephrased the termination problem as a second-order satisfaction problem and made 
use of results in genetic programming to efficiently solve it. We have also investigated the effects of genetic operators on the search space for ranking functions and computed theoretical 
bounds on the convergence time ...
\item We present a formulation that allows us to prove termination without an expensive reachability check (as we don't use Ramsey-based termination arguments).  In particular,
we only need a bounded model checker that performs a single unwinding of a loop.
%% \item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
%% multiple loops.

\item We iteratively construct a termination argument as a pair of a ranking function and supporting invariant through a single refinement loop.
This is opposed to tools that alternate 
between calls to a safety prover and calls to a ranking function synthesis tool \cite{}.

\item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.

\end{itemize}

@@

Proving program termination implies the simultaneous search for a termination argument and a supporting invariant.
For this purpose, several tools alternate 
between calls to a safety prover and calls to a ranking function synthesis tool \cite{}. 
In order to share more information about the state of the termination proofs between the two tools, 
Brockschmidt et al. use a single representation of the state of the termination proof search that both tools operate over, 
resulting in performance improvements \cite{DBLP:conf/cav/BrockschmidtCF13}.
In our setting, both the ranking function and the supporting invariant are iteratively constructed through a single refinement loop.

%In our setting, a candidate termination argument is iteratively constructed. 
%A candidate termination argument is composed out of a ranking function 
%and a supporting invariant.

%% The safety prover proves or disproves the
%% validity of the current argument via the search for invariants. Refinement of the
%% current termination argument is performed using the output of a rank function
%% synthesis tool when applied to counterexamples found by the safety prover.

@@

When proving program termination we are simultaneously solving two problems:
the search for a termination argument, and the search for a supporting invariant.
In our setting, a candidate termination argument is iteratively constructed. 
A candidate termination argument is composed out of a ranking function 
and a supporting invariant.

The program in Figure~\ref{fig:motivation.i} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
\begin{verbatim}
R(x, y, z) = (x < y, z)
\end{verbatim}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
  y = 1;

  while (x > 0) {
    x = x-y;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.h}
\end{subfigure}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0 && y > 0 && z > 0) {
    if (y > x) {
      y = z;
      x = nondet();
      z = x - 1;
    } else {
      z = z - 1;
      x = nondet();
      y = x - 1;
    }
}
\end{lstlisting}
\caption{}
 \label{fig:motivation.i}
\end{subfigure}

@@

\section{Preliminaries}
Given a transition system, %with state space $X$ and transition relation $T \subseteq X \times X$, 
we first formalise its termination argument as a ranking function (Section~\ref{sec:ranking.functions}). 
Subsequently, we discus bit-vector semantics and illustrate 
differences between machine arithmetic and integer arithmetic that
make the abstraction of bit-vectors to mathematical integers unsound (Section~\ref{sec:machine.arith}).

\subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
\todo{linear program}

\begin{definition}[Unconditional termination]
A transition system with state space $X$ and transition relation $T \subseteq X \times X$
is said to be \emph{unconditionally terminating} if there is no infinite sequence
$\vec{x_1}, \vec{x_2}, \ldots$ with $\forall i . T(\vec{x_i}, \vec{x_{i+1}})$.  
\end{definition}

We can prove that $T$ is
unconditionally terminating by finding a ranking function.

\begin{definition}[Ranking function]
Given a transition system with state space $X$ and transition relation $T \subseteq X \times X$,
a ranking function is an injective function $R: X \to Y$ where
$Y$ is well-founded and $R$ is monotonically decreasing with respect to $T$.  That is
to say:
$$\forall \vec{x}, \vec{x'} . T(\vec{x}, \vec{x'}) \Rightarrow R(\vec{x}) < R(\vec{x'})$$
\end{definition}

@@

%% \subsection{Disjunctive Well-Foundedness}\label{sec:dwf}
%% Terminator uses disjunctive well foundedness arguments, which can be found
%% piece by piece, but require reasoning about the transitive closure of a transition
%% relation.  Because we build a monolithic well foundedness argument, we
%% do not need to reason about the full transitive closure, and can instead do our
%% reasoning over a single step of the transition relation.  This means that we
%% can get away with very simple model checking methods, rather than requiring
%% a full-blown safety prover.

@@

The existence of a ranking function for a single loop is equivalent to the satisfiability
of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
$R$ is a ranking function that proves loop $L$ terminates.
